Simulating High-Dimensional Multivariate Data

### 2020-12-17 050-examples.Rmd
4th draft. Address WWP ch050

# RNA-seq data example applications  {#examples}

@@

This section demonstrates how to simulate multivariate data using `bigsimr`, aiming to replicate the structure of high-dimensional dependent count data.
In an illustration of our proposed methodology applied to real data, we seek to simulate RNA-sequencing data by producing simulated random vectors with assumed marginal distributions with estimated parameters that mimic the observed data and its generating process.
We seek scaleable realistic multivariate simulation to enable Monte Carlo (MC) methods for these data.
Modeling RNA-seq using multivariate probabilities distributions is natural as inter-gene correlation is an inherent part of biological processes.
Yet many models do not account for this, leading to major disruptions to the operating characteristics of statistical estimation, testing, and prediction.
See @Efron2012 for a detailed discussion with related methods and see @Wu2012b, @Schissler2018, @Schissler2019 for applied examples.
The following subsections apply `bigsimr`'s methods to real RNA-seq data, including replicated an estimated parametric structure, MC probability estimation, and MC evaluation of correlation estimation efficiency.

@@

Our first goal is to replicate the structure of the TCGA BRCA RNA-seq data set (see [Background](background)).
Ultimately, we will simulate $B=10,000$ random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with $d=`r d`$.
Often researchers posit a negative binomial (NB) model as RNA-seq counts are often over-dispersed that a Poisson model would suggest.
All $d$ selected genes exhibit over-dispersion (data not shown) and, so, we proceed to estimate the NB parameters $(r_i, p_i), i=1,\ldots,d$ to determine the target marginal PMFs $f_i$ (via method of moments).
To complete specification of the simulation algorithm inputs, we'll estimate the Spearman correlation matrix ${ \bf R}_{Spearman}$ to characterize dependency.

@@

With our goal in mind, we first estimate the desired correlation matrix using the fast implementation provided by `bigsimr`:

@@

Next, we estimate the marginal parameters. 
We use method of moments (MoM) to estimate the marginal parameters for the multivariate negative binomial model.
The marginal distributions are from the same probability family (NB) yet are heterogeneous in terms of the parameters probability and size $(p_i, n_i)$ for $i,\ldots,d$.
The functions below help achieve this estimation and specifying the inputs for use in `bigsimr::rvec`.

@@

Now we estimate the marginal parameters using the built-in `apply`function:

@@

Notably, the marginal NB probabilities $\hat{p}_i's$ are small --- ranging in $[`r min(probs)` , `r max(probs)`]$.
This gives rise to highly variable counts and, typically, less restriction on potential pairwise correlation pairs.
Once the functions are defined/executed to complete marginal estimation, we specify targets and generate the desired random vectors using `rvec`:

@@

Figure \@ref(fig:ch050-simDataFig) displays the simulated counts and pairwise relationships for our example genes in Table \@ref(tab:ch010-realDataTab).
Simulated counts roughly mimic the observed data but with a smoother appearance due to the assumed parameter form and with less extreme points then the observed data (c.f. Figure \@ref(fig:ch010-realDataFig).)

@@

Figure \@ref(fig:ch050-figBRCA) displays the aggregated results of our simulation by comparing the specified target parameter (horizontal axes) with the corresponding quantities estimated from the simulated data (vertical axes).
The evaluation shows that the simulated counts approximately match the target parameters and exhibit the full range of estimated correlation from the data.
Utilizing 15 CPU threads in a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, the simulation completed just over of 2 minutes.

@@

where ${\bf Y^{(b)} }$ is the $b^{th}$ simulated vector in a total of $B$ simulation replicates and $I ( )$ is the indicator function.
For example, we can estimate from our $B=10,000$ simulated vectors the probability that all genes are expressed (i.e., ${\bf y}_i \geq 1, \forall \; i$) is $0.1708$.

```{r ch050-densityEvaluation, echo=TRUE, eval=TRUE, cache=F}
d <- ncol(sim_nbinom)
B <- nrow(sim_nbinom)
threshold <- rep( 1, d)
mean(apply( sim_nbinom,  1,
           function(X, y0=threshold) {
               all( X > y0) }
           ))
```

@@

MC methods are routinely used in many statistical inferential tasks including estimation, hypothesis testing, error rates, and empirical interval coverage rates.
For an concise introduction to these methods, see @Rizzo2007, Ch. 6.
To conclude the example applications, we demonstrate how `bigsimr` can be used evaluate estimation efficiency.
In particular, we'd like to assess the error in our correlation estimation above.
We used a conventional method, based on classical statistical theory. 
Yet this method was not designed for high-dimensional data.
Indeed, high-dimensional covariance estimation (and precision matrices) is an active area of statistical science (see, for example, [@Won2013g; @VanWieringen2016].

@@

This distribution could be compared to high-dimensional designed covariance estimators to guide to help decide whether the additional complexity and computation time are warranted.

>>>>

# RNA-seq data applications  {#examples}

@@

This section demonstrates how to simulate multivariate data using `bigsimr`, aiming to replicate the structure of high-dimensional dependent count data.
In an illustration of our proposed methodology, we seek to simulate RNA-sequencing data by producing simulated random vectors with assumed marginal distributions and with estimated parameters that mimic the observed data and its generating process.
We seek scaleable realistic multivariate simulation to enable Monte Carlo (MC) methods for these data.
Modeling RNA-seq using multivariate probabilities distributions is natural as inter-gene correlation is an inherent part of biological processes [@Wang2009b].
Yet many models do not account for this, leading to major disruptions to the operating characteristics of statistical estimation, testing, and prediction.
See @Efron2012 for a detailed discussion with related methods and @Wu2012b, @Schissler2018, @Schissler2019 for applied examples.
The following subsections apply `bigsimr`'s methods to real RNA-seq data, including replicating an estimated parametric structure, MC probability estimation, and MC evaluation of correlation estimation efficiency.

@@

Our initial goal is to replicate the structure of the TCGA BRCA RNA-seq data set (see [Background](background)).
Ultimately, we will simulate $B=10,000$ random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with $d=`r d`$.
Often researchers posit a negative binomial (NB) model as RNA-seq counts are often over-dispersed.
All $d$ selected genes exhibit over-dispersion (data not shown) and, so, we proceed to estimate the NB parameters $(r_i, p_i), i=1,\ldots,d$ to determine the target marginal PMFs $f_i$ (via the method of moments).
To complete specification of the simulation algorithm inputs, we estimate the Spearman correlation matrix ${ \bf R}_{Spearman}$ to characterize dependency.

@@

With this goal in mind, we first estimate the desired correlation matrix using the fast implementation provided by `bigsimr`:

@@

Next, we estimate the marginal parameters. 
We use the method of moments (MoM) to estimate the marginal parameters for the multivariate negative binomial model.
The marginal distributions are from the same probability family (NB) yet are heterogeneous in terms of the parameters probability and size $(p_i, n_i)$ for $i,\ldots,d$.
The functions below help achieve this estimation and specifying the inputs for use in `bigsimr::rvec`.

@@

Now we estimate the marginal parameters using the built-in `apply` function:

@@

Notably, the marginal NB probabilities $\hat{p}_i's$ are small --- ranging in the interval $[`r min(probs)` , `r max(probs)`]$.
This gives rise to highly variable counts and, typically, less restriction on potential pairwise correlation pairs.
Once the functions are defined/executed to complete marginal estimation, we specify targets and generate the desired random vectors using `rvec`:

@@

Figure \@ref(fig:ch050-simDataFig) displays the simulated counts and pairwise relationships for our example genes from Table \@ref(tab:ch010-realDataTab).
Simulated counts roughly mimic the observed data but with a smoother appearance due to the assumed parameter form and with less extreme points then the observed data in Figure \@ref(fig:ch010-realDataFig).

@@

Figure \@ref(fig:ch050-figBRCA) displays the aggregated results of our simulation by comparing the specified target parameter (horizontal axes) with the corresponding quantities estimated from the simulated data (vertical axes).
The evaluation shows that the simulated counts approximately match the target parameters and exhibit the full range of estimated correlation from the data.
Utilizing 15 CPU threads in a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, the simulation completed in just over 2 minutes.

@@

```{r ch050-densityEvaluation, include=FALSE, eval=TRUE, cache=F}
d <- ncol(sim_nbinom)
B <- nrow(sim_nbinom)
threshold <- rep( 1, d)
pHat <- mean(apply( sim_nbinom,  1,
           function(X, y0=threshold) {
               all( X >= y0) }
           ))
```

where ${\bf Y^{(b)} }$ is the $b^{th}$ simulated vector from a total of $B$ simulation replicates and $I ( )$ is the indicator function.
For example, we can estimate from our $B=10,000$ simulated vectors that the probability of all genes expressing (i.e., ${\bf y}_i \geq 1, \forall \; i$) is $`r pHat`$.

```{r ch050-densityEvaluationECHO, echo=TRUE, eval=TRUE, cache=F}
d <- ncol(sim_nbinom)
B <- nrow(sim_nbinom)
threshold <- rep( 1, d)
mean(apply( sim_nbinom,  1,
           function(X, y0=threshold) {
               all( X >= y0) }
           ))
```

@@

MC methods are routinely used in many statistical inferential tasks including estimation, hypothesis testing, error rates, and empirical interval coverage rates.
To conclude the example applications, we demonstrate how `bigsimr` can be used evaluate estimation efficiency.
In particular, we wish to assess the error in our correlation estimation above.
We used a conventional method, based on classical statistical theory, which unfortunately, was not designed for high-dimensional data.
Indeed, high-dimensional covariance estimation (and precision matrices) is an active area of statistical science (.e.g., @Won2013g and @VanWieringen2016).

@@

This distribution could be compared to high-dimensional designed covariance estimators for guidance on whether the additional complexity and computation time are warranted.
